\section{Appendix D: In Process and Wish List Items for Future Versions}

\begin{center}
	\begin{longtable}{p{2cm} p{3cm} p{10cm}}
		Category & Item & Description\\
		\hline
		\endfirsthead
				
		Category & Item & Description\\
		\hline
		\endhead
				
		\hline
		\endfoot
				
		\endlastfoot
		
		In process 
		  &	Environmental survey of a deviation vector & Environmental survey data can be related to f(a deviation vector), but needs more Q\_link functions.\\
		  \\
		  & Retrospective & Make blocks work when doing retrospective analyses. \\
		
		\hline
		Wishlist & &
			 Tag-recapture major revamp\\\\
	     & & Area specific spawner-recruitment \\\\
	     & & Add automatic set-up for size selectivity option 6 and age selectivity option 17 given the data (comparable to the current capacity for cubic spline selectivity).\\\\
	     & & More error checking on read of empirical weight-at-age and composition data. \\\\
	     & & New version of age-specific K needed because current version is inefficient.\\\\
	     & & Mean size in plus group uses a fixed erosion factor of 0.2; should be context specific. \\\\
	     & & Consider Tim Miller's state space model approach. \\\\
	     & & Add other options to CV growth patter for log(SD). \\\\
	     & & Add measure of auto-correlation in composition and in survey residuals.\\\\
	     & & Add SD report for survey expected values. \\\\
	     & & Add calculation of Francis composition weighting method. \\\\
	     & & Year-specific MSY.	     	   \\	  	  				
		\hline
	\end{longtable}
\end{center}

